Numerical investigation of viscous flow instabilities through heterogeneous porous media. Advantages and limitations of unstructured and adaptive meshes in 2D and 3D domains under different mobility scenarios.

### 2017-12-07 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

%%%%%%%%%%%%%
\subsection{3-D channel flows} \label{section:results_3D} 
In large scale geological formations, preferential flow may result in fast polutant transport (advection-diffusion-dispersion) after industrial spillage (therefore leading to contamination of groundwater) and low hydrocarbon recovery rates (in oil/gas exploration fields).  In this section, 3D simulations are introduced to investigate preferential flow pathways through semi-pervious and impervious geological layers.

@@

\medskip
In Sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}, a $2$D domain was used to describe one or more finger formations. These flow instabilities were triggerded from the permeability differences in the domain. Using a $3D$ domain, the difference between a flow instability and the preferential flow path that may include smaller fingers is more obvious as the latter may include finger formations. Figure \ref{fig:3DChannel_sat} describes the flow path in various instances of time showing the progression of the saturation front that now becomes more distinct. Heterogeneity is determine by the spatial distribution of porosity ($\phi =0.20$ in the current case) and permeability.

\medskip 
The progress of the saturation front and formation and growth of fingers in time and space can be seen in more details in isosurfaces shown in Fig.~\ref{fig:3DChannel_Isosurf}. However due to relatively low mesh resolution, in particular above and below the central channel region, capture of fingers (formation and growth) was very limited. In figures \ref{fig:3DChannel_sat_adapt}(a)-(f) an unstructured and adaptive mesh has been introduced in order to achieve higher resolution and focus the computational effort where is needed in he domain. For the same time-steps an inner view on the domain provides with a more detailed description of not only the flow path itself as well as the minor fingers that may be created. At this point resolution plays an important role on capturing these patterns in detail and is also a limititation that needs to be overcome.

\medskip
A quick review to the mechanisms of viscous fingering (\ie shielding, spreading and tip-splitting, \citet{homsy_1987}) is required. In all the $2$D previous cases (sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}), as interfacial surface tension (IFT) becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability. Through \textit{shielding} mechanism, one of these fingers will eventually outgrow the other and, will then spread (\textit{spreading}) to occupy the appropriate width of the domain ahead of it. Following this process, it reaches a width that is again unstable and the pattern repeats. So far shielding and spreading were dominating the flow but, using the current $3$D domain and by applying two types of meshes, despite the resolution restrictions, the conditions and the mechanics of tip-splitting along a front can be qualitatively understood.

>>>>

%%%%%%%%%%%%%
\subsection{3-D channel flows} \label{section:results_3D} 
In geological formations, preferential flow may result in fast polutant transport (advection-diffusion-dispersion) after industrial spillage (therefore leading to contamination of groundwater) or low hydrocarbon recovery rates (in oil/gas exploration fields). In this Section, 3D simulations are introduced to investigate preferential flow pathways through semi-pervious and impervious geological layers.

@@

In Sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}, finger's formation, growth and braching were triggerded by domain heterogeneity represented by permeability differences. Figure~\ref{fig:3DChannel_sat} shows preferential flow pathways at various instances of time. 
%In 3D simulations, the difference between a flow instability and the preferential flow path that may include smaller fingers is more obvious as the latter may include finger formations. Figure \ref{fig:3DChannel_sat} describes the flow path in various instances of time showing the progression of the saturation front that now becomes more distinct. Heterogeneity is determine by the spatial distribution of porosity ($\phi =0.20$ in the current case) and permeability. 
The progress of the saturation front and formation and growth of fingers in time and space can be seen in more details in isosurfaces shown in Fig.~\ref{fig:3DChannel_Isosurf}. However, due to relatively low mesh resolution, in particular above and below the central channel region, capture of fingers (formation and growth) was very limited. A numerical simulation of the same problem (\ie same geometry, initial mesh, initial and boundary conditions) was also performed using an adaptive mesh, Fig.~\ref{fig:3DChannel_sat_adapt}. For the same time-steps an inner view on the domain provides with a more detailed description of not only the flow path itself as well as the minor fingers that may be created. At this point resolution plays an important role on capturing these patterns in detail and is also a limititation that needs to be overcome.

\medskip
A quick review to the mechanisms of viscous fingering (\ie shielding, spreading and tip-splitting, \citet{homsy_1987}) is required. In all the $2$D previous cases (sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}), as interfacial surface tension (IFT) becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability. Through \textit{shielding} mechanism, one of these fingers will eventually outgrow the other and, will then spread (\textit{spreading}) to occupy the appropriate width of the domain ahead of it. Following this process, it reaches a width that is again unstable and the pattern repeats. So far shielding and spreading were dominating the flow but, using the current $3$D domain and by applying two types of meshes, despite the resolution restrictions, the conditions and the mechanics of tip-splitting along a front can be qualitatively understood.\red{WHAT IS THE LINKAGE BETWEEN THIS PARAGRAPH AND THE REMAINING OF THE SECTION?}

### 2017-12-07 Papers/ViscousInstabilities/article_figure1.tex
updates

%%%
%%% FIGURE XXXXXX
%%%
\begin{figure}[ht]
\vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=\textwidth]{./Pics1/3D_Channel/Test_SlowNew_MeshPermeability.pdf} }}
\caption{3D channel flow: permeability distribution. The domain contains 243056 \PN[1]{1} element-pairs.}
\label{fig:3DChannel_PermMesh}
\end{figure}
\clearpage

%%%%
%%%%  FIGURE
%%%%
%\begin{landscape}
\begin{figure}[ht] 
\vbox{
\hbox{\hspace{0.5cm}
\includegraphics[width=1.0\textwidth]{./Pics1/3D_Channel/3D_channel_darcy_vel_planes_490_1_1.pdf} 
}
\vspace{-12.0cm}
\hbox{\hspace{6.0cm} (a)     
}
\hbox{\hspace{0.5cm}
\includegraphics[width=1.0\textwidth]{./Pics1/3D_Channel/3D_channel_darcy_vel_planes_30_1_1.pdf}
}
\vspace{-12.0cm}
\hbox{\hspace{6.0cm} (b)      
}
}     
\caption{3D channel flow: XY and YZ planes showing wetting fluid phase Darcy velocity vectors overlapped with wetting fluid saturation at (a) t=$0.29$s and (b)t=$4.3$s.}
\label{fig:3DChannel_planes}
\end{figure}
%\end{landscape}
\clearpage

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
\begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.6\textwidth]{./Pics1/3D_Channel/3D_channel_sat_30.pdf} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_60.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_90.pdf} } 
      \hbox{\hspace{2.cm} (a) t=0.29s \hspace{5.cm} (b) t=0.56s \hspace{4.cm} (c) t=0.82s}
      \vspace{0.5cm}
      \hbox{\hspace{.5cm} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_110.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_150.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_490.pdf} }
      \hbox{\hspace{2.cm} (d) t=1.00s \hspace{5.cm} (e) t=1.35s\hspace{4.cm} (f) t=3.50s  } }
\caption{3D channel flow: saturation front evolving in time and space with preferential flow pathways through 3.50 seconds of numerical simulations. Preferential flow pathway can be readily noticed in these frames, mirroring permeability distribution. The domain contains 368513 \PN[1]{1} elements.}
\label{fig:3DChannel_sat}
\end{figure}
\end{landscape}
\clearpage

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
  \begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.55\textwidth, height=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D30c.pdf} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D60c.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D90c.pdf} }
      \hbox{\hspace{2.0cm} (a) t=0.29s \hspace{6.cm} (b) t=0.56s \hspace{4.cm} (c) t=0.82s}
      \vspace{0.5cm}
      \hbox{\hspace{.5cm} \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D110c.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D150c.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D400c.pdf} }
      \hbox{\hspace{3.cm} (d) t=1.00s \hspace{3.cm} (e) t=1.35s\hspace{4.cm} (f) t=3.50s }}
\caption{3D channel flow: Isosurfaces for wetted phase saturatioh ranging from 0.45 and 0.60 at the same instants of time of Fig.~\ref{fig:3DChannel_sat}. Preferential flow pathway can be readily noticed in (b) and (f). Fingers' formation and growth can be clearly noticed in (b)-(e). The domain contains 368513 \PN[1]{1} elements.}
\label{fig:3DChannel_Isosurf}
\end{figure}
\end{landscape}
\clearpage

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
  \begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.5\textwidth, height=.45\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_30.pdf} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_60.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_90.pdf} }
      \hbox{\hspace{3.0cm} (a) t=0.29s \hspace{4.cm} (b) t=0.56s \hspace{4.cm} (c) t=0.82s}
      \vspace{1.5cm}
      \hbox{\hspace{.5cm} \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_110.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_150.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_288.pdf} }
      \hbox{\hspace{3.cm} (d) t=1.00s \hspace{3.cm} (e) t=1.35s\hspace{4.cm} (f) t=3.50s }}
\caption{For the same time steps as these are shown in figure \ref{fig:3DChannel_sat} there is a description of the same flow path but now using an adaptive and unstructured mesh. The aim was to describe in greater detail the finger patterns, as the flow develops but also address the importance of a resolution, in this case the type of mesh that is used to capture these phenomena.}
\label{fig:3DChannel_sat_adapt}
\end{figure}
\end{landscape}
\clearpage

%%%%
%%%%  FIGURE
%%%%
%\begin{landscape}
\begin{figure}[ht] 
\vbox{\vspace{-3.5cm}
\vbox{\hspace{2.5cm}
\includegraphics[width=0.75\textwidth]{./Pics1/3D_Channel/3D_channel_sat_end.pdf} 
}
\vspace{0.0cm}
\hbox{\hspace{2.0cm} (a) fixed unstructured mesh of 322493 elements    
}
\hbox{\hspace{2.5cm}
\includegraphics[width=0.75\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_end.pdf}
}
\vspace{0.0cm}
\hbox{\hspace{-2.0cm} (b) adaptive and unstructured mesh with a maximum of 1242202 elements at the end of the simulation    
}
}     
\caption{This is a qualitative comparison at t=end of simulation, for bot the fixed and unstructured mesh used in the 3D domain.}
\end{figure}
%\end{landscape}
\clearpage

@@

%%%%
%%%%  FIGURE
%%%%
%\begin{landscape}
  \begin{figure}[ht] 
    \vbox{\vspace{0.cm}
        \hbox{\hspace{-3.cm}
            \includegraphics[width=.75\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_0_ContourPlotSlice_b} \hspace{-.5cm}
            \includegraphics[width=.70\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_20_ContourPlotSlice_b}}
        \hbox{\hspace{-3.cm}
            \includegraphics[width=.70\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_60_ContourPlotSlice_b}
            \includegraphics[width=.70\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_120_ContourPlotSlice_b}}
        \hbox{\hspace{-3.cm}
            \includegraphics[width=.70\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_200_ContourPlotSlice_b}
            \includegraphics[width=.70\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_288_ContourPlotSlice_b}}
}
\caption{3D Channel: Text to come \red{should i mention something here? something qualitative}}
\end{figure}
%\end{landscape}
\clearpage

>>>>

%%%
%%% FIGURE XXXXXX
%%%
\begin{figure}[ht]
\vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=\textwidth]{./Pics1/3D_Channel/Test_SlowNew_MeshPermeability.pdf} }}
\caption{3D channel flow (fixed mesh): permeability distribution. The domain contains 243056 \PN[1]{1} element-pairs.}
\label{fig:3DChannel_PermMesh}
\end{figure}
\clearpage

%%%%
%%%%  FIGURE
%%%%
%\begin{landscape}
\begin{figure}[ht] 
\vbox{
\hbox{\hspace{0.5cm}
\includegraphics[width=1.0\textwidth]{./Pics1/3D_Channel/3D_channel_darcy_vel_planes_490_1_1.pdf} 
}
\vspace{-12.0cm}
\hbox{\hspace{6.0cm} (a)     
}
\hbox{\hspace{0.5cm}
\includegraphics[width=1.0\textwidth]{./Pics1/3D_Channel/3D_channel_darcy_vel_planes_30_1_1.pdf}
}
\vspace{-12.0cm}
\hbox{\hspace{6.0cm} (b)      
}
}     
\caption{3D channel flow (fixed mesh): XY and YZ planes showing wetting fluid phase Darcy velocity vectors overlapped with wetting fluid saturation at (a) t=$0.29$s and (b)t=$4.3$s.}
\label{fig:3DChannel_planes}
\end{figure}
%\end{landscape}
\clearpage

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
\begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.6\textwidth]{./Pics1/3D_Channel/3D_channel_sat_30.pdf} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_60.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_90.pdf} } 
      \hbox{\hspace{2.cm} (a) t=0.29s \hspace{5.cm} (b) t=0.56s \hspace{4.cm} (c) t=0.82s}
      \vspace{0.5cm}
      \hbox{\hspace{.5cm} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_110.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_150.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_490.pdf} }
      \hbox{\hspace{2.cm} (d) t=1.00s \hspace{5.cm} (e) t=1.35s\hspace{4.cm} (f) t=3.50s  } }
\caption{3D channel flow (fixed mesh): saturation front evolving in time and space with preferential flow pathways through 3.50 seconds of numerical simulations. Preferential flow pathway can be readily noticed in these frames, mirroring permeability distribution. The domain contains 368513 \PN[1]{1} elements.}
\label{fig:3DChannel_sat}
\end{figure}
\end{landscape}
\clearpage

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
  \begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.55\textwidth, height=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D30c.pdf} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D60c.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D90c.pdf} }
      \hbox{\hspace{2.0cm} (a) t=0.29s \hspace{6.cm} (b) t=0.56s \hspace{4.cm} (c) t=0.82s}
      \vspace{0.5cm}
      \hbox{\hspace{.5cm} \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D110c.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D150c.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D400c.pdf} }
      \hbox{\hspace{3.cm} (d) t=1.00s \hspace{3.cm} (e) t=1.35s\hspace{4.cm} (f) t=3.50s }}
\caption{3D channel flow (fixed mesh): Isosurfaces for wetted phase saturatioh ranging from 0.45 and 0.60 at the same instants of time of Fig.~\ref{fig:3DChannel_sat}. Preferential flow pathway can be readily noticed in (b) and (f). Fingers' formation and growth can be clearly noticed in (b)-(e). The domain contains 368513 \PN[1]{1} elements.}
\label{fig:3DChannel_Isosurf}
\end{figure}
\end{landscape}
\clearpage

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
  \begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.5\textwidth, height=.45\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_30.pdf} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_60.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_90.pdf} }
      \hbox{\hspace{3.0cm} (a) t=0.29s \hspace{4.cm} (b) t=0.56s \hspace{4.cm} (c) t=0.82s}
      \vspace{1.5cm}
      \hbox{\hspace{.5cm} \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_110.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_150.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_288.pdf} }
      \hbox{\hspace{3.cm} (d) t=1.00s \hspace{3.cm} (e) t=1.35s\hspace{4.cm} (f) t=3.50s }}
\caption{3D channel flow (adaptive mesh): numerical simulation was performed with the smae boundary and initial conditions as in Figs.~\ref{fig:3DChannel_PermMesh}-\ref{fig:3DChannel_Isosurf} with an adaptive mesh. Flow pathway is very similar to the one shown in Fig.~\ref{fig:3DChannel_sat}.}
\label{fig:3DChannel_sat_adapt}
\end{figure}
\end{landscape}
\clearpage

\begin{comment}
%%%%
%%%%  FIGURE
%%%%
%\begin{landscape}
\begin{figure}[ht] 
\vbox{\vspace{-3.5cm}
\vbox{\hspace{2.5cm}
\includegraphics[width=0.75\textwidth]{./Pics1/3D_Channel/3D_channel_sat_end.pdf} 
}
\vspace{0.0cm}
\hbox{\hspace{2.0cm} (a) fixed unstructured mesh of 322493 elements    
}
\hbox{\hspace{2.5cm}
\includegraphics[width=0.75\textwidth]{./Pics1/3D_Channel/3D_channel_sat_adapt_end.pdf}
}
\vspace{0.0cm}
\hbox{\hspace{-2.0cm} (b) adaptive and unstructured mesh with a maximum of 1242202 elements at the end of the simulation    
}
}     
\caption{This is a qualitative comparison at t=end of simulation, for bot the fixed and unstructured mesh used in the 3D domain.}
\end{figure}
%\end{landscape}
\clearpage
\end{comment}

%%%%
%%%%  FIGURE
%%%%
%\begin{landscape}
  \begin{figure}[ht] 
    \vbox{\vspace{0.cm}
        \hbox{\hspace{-3.cm}
            \includegraphics[width=.75\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_0_ContourPlotSlice_b} \hspace{-.5cm}
            \includegraphics[width=.70\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_20_ContourPlotSlice_b}}
        \hbox{\hspace{2cm}(a)\hspace{9cm}(b)}
        \hbox{\hspace{-3.cm}
            \includegraphics[width=.70\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_60_ContourPlotSlice_b}
            \includegraphics[width=.70\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_120_ContourPlotSlice_b}}
        \hbox{\hspace{2cm}(c)\hspace{9cm}(d)}
        \hbox{\hspace{-3.cm}
            \includegraphics[width=.70\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_200_ContourPlotSlice_b}
            \includegraphics[width=.70\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_288_ContourPlotSlice_b}}
        \hbox{\hspace{2cm}(e)\hspace{9cm}(f)}
}
    \caption{3D Channel (adaptive mesh):  XY and YZ planes showing wetting fluid saturation overlapped with Darcy velocity vectors at  (a) 0.00, (b) 0.09, (c) 0.71, (d) 1.75, (e) 2.47 and (f) 3.78s. Colour scheme for saturation profile is the same as used in Fig.~\ref{fig:3DChannel_sat_adapt}.}
\end{figure}
%\end{landscape}
  \clearpage

@@
